---
title: "Fixed effects models"
sidebar: false
eval: false
---

- **You can download a template file for this activity [here](templates/13-fe.qmd).**
- Slides for today are available [here](https://docs.google.com/presentation/d/1GFgEghgpCxKmufScPU5W3Lci-0g694qYbGkAs1uVUUI/edit?usp=sharing).

# Exploration

```{r}
library(tidyverse)
library(fixest)
library(lme4)
library(broom)
library(causaldata)
```


## Gapminder data

We have yearly life expectancy data for all countries of the world from the [Gapminder Institute](https://www.gapminder.org/). We also have yearly GDP data.

Let's first load the Gapminder data:

```{r}
gm <- causaldata::gapminder
```

### Visualization

The first visualization shows the relationship between log(GDP) and life expectancy that would be estimated using the overall data without capitalizing on looking at variation within a geographical unit.

The second visualization shows trends by continent.

**Questions:**

- Which approach to exploring the relationship do you think is better and why?
- How would our inferences about the relationship between log(GDP) and life expectancy differ between the approaches?
- Now consider a generic situation with outcome Y, quantitative treatment A, and grouping variable G. Draw a scatterplot that depicts a situation in which the within-group relationship between Y and A shows a negative relationship but the overall relationship (ignoring group) shows a positive relationship.

```{r message=FALSE}
# One overall trend line
ggplot(gm, aes(x = log(gdpPercap), y = lifeExp, color = continent)) +
    geom_point(alpha = 0.1) +
    geom_smooth(aes(x = log(gdpPercap), y = lifeExp), data = gm, color = "red", size = 2, method = "lm") +
    coord_cartesian(xlim = c(5,12), ylim = c(20,95)) +
    theme_classic()

# Overall trend line + continent-specific trends
ggplot(gm, aes(x = log(gdpPercap), y = lifeExp, color = continent)) +
    geom_point(alpha = 0.1) +
    geom_smooth(aes(x = log(gdpPercap), y = lifeExp), data = gm, color = "red", size = 2, method = "lm") +
    geom_smooth(method = "lm", se = FALSE) +
    coord_cartesian(xlim = c(5,12), ylim = c(20,95)) +
    theme_classic()
```

### Fixed effects in `fixest`

The `fixest` package provides a fast way of estimating fixed effects models. The coefficient estimates are the same as from `lm()`, but it provides ways of adjusting the standard errors for correlation within a group.

Below we use `feols()` function from `fixest` to fit a fixed effects model with country-specific indicator variables.

```{r}
# Model formula: Y ~ A + time invariant covariates + time fixed covariates | variables that we want indicators for
gm_mod_fe <- feols(lifeExp ~ log(gdpPercap) | country, data = gm)
summary(gm_mod_fe, cluster = ~ country)
```

- Interpretation: Within a given country, every one unit increase in log(GDP) is expected to increase life expectancy by 9.77 years.





We can also have more than one set of fixed effects. It is common to have multiple observations for a unit over time, so we can have fixed effects for both unit and time (indicator variables for each unit and indicators for each time period). This is called a **two-way fixed effects (TWFE)** model.

```{r}
gm_mod_twfe <- feols(lifeExp ~ log(gdpPercap) | country + year, data = gm)
summary(gm_mod_twfe, cluster = ~ country)
```

- Interpretation: Within a given country and year, every one unit increase in log(GDP) is expected to increase life expectancy by 1.45 years.


For comparison purposes, let's look at the results from using `lm()`.

- We see the same coefficient estimates.
- We have lower standard errors with `lm()` because assuming all observations are independent is assuming that we have a lot more information than if some observations are correlated (as within a country).

```{r}
gm_mod_fe_lm <- lm(lifeExp ~ log(gdpPercap) + country, data = gm)
tidy(gm_mod_fe_lm) %>% filter(str_detect(term, "gdp"))

gm_mod_twfe_lm <- lm(lifeExp ~ log(gdpPercap) + country + factor(year), data = gm)
tidy(gm_mod_twfe_lm) %>% filter(str_detect(term, "gdp"))
```


### Random effects using `lmer`

Another approach to fitting fixed effects models is to fit what are known as **random effects models**. Here we don't estimate

A standard fixed effects model with one set of fixed effects for units can be written as:

$$
Y_{it} = \beta_0 + b_i + \beta X_{it} + \epsilon_{it}
$$

where 

- $b_i$ are independent random variables following a $\text{Normal}(0, \sigma_b^2)$ distribution
- $X_{it}$ is a set of covariates
- $\epsilon_{it}$ are independent random variables following a $\text{Normal}(0, \sigma_e^2)$ distribution.

Key points:

- Each observation from unit $i$ share the same $b_i$. This induces correlation between repeated observations from the same unit.
- The individual $b_i$'s are not estimated. Only the variance of the normal distribution $\sigma_b^2$ is estimated. Estimating fewer parameters increases statistical power.

We can use the `lmer()` function in the `lme4` package to estimate this random effects model. (In the statistical literature, such a model is called a **mixed effects model**.)

```{r}
# Model formula for one-way FEs: Y ~ (1|unit_var) + covariates
# Model formula for two-way FEs: Y ~ (1|unit_var) + (1|time_var) + covariates
gm_mod_re <- lmer(lifeExp ~ (1|country) + (1|year) + log(gdpPercap), data = gm)
summary(gm_mod_re)
```

- Interpretation: Within a given country and year, a one unit increase in log(GDP) is expected to increase life expectancy by 2.53 years.

Note that this estimate is different from our fixed effects models. Why?

**Important note:** The statistical formulation of a random effects model assumes that $b_i$ (the unit specific effects) are independent of (and thus uncorrelated with) the covariates in the model.

- In other words, this says that all the time-fixed covariates that go into a unit-specific effect $b_i$ are unrelated to the other variables in our model.
    - In causal inference, the key other variable in our model is the treatment variable.
    - So in a random effects model, the unit-specific factors are assumed to be uncorrelated with other variables in our model, including treatment.
- In this context, time-fixed covariates that likely affect life expectancy include country geography, climate, environment, ...
    - In our random effects model, these covariates are assumed to be uncorrelated with GDP. This seems unlikely.

The [last section](https://www.theeffectbook.net/ch-FixedEffects.html#advanced-random-effects) of our textbook chapter discusses alterations to standard random effects models that can handle this.





## Data on school expenditures and student outcomes

**Research question:** What is the effect of expenditures on math outcomes for 4th graders?

We have data across many school districts over multiple years. Focus on the following variables:

- `distid`: the district identifier
- `year`: the year the data is from
- `math4`: the percentage of 4th grade students who are "satisfactory" or better in math (outcome)
- `expp`: expenditure per pupil (treatment)
- `lunch`: the percentage of students eligible for free lunch (time-varying confounder as a proxy for district socioeconomic status)

```{r}
school <- read_csv("https://raw.githubusercontent.com/NickCH-K/TheEffectAssignments/refs/heads/main/mathpnl.csv")

# Make distid and year categorical
school <- school %>% 
    mutate(distid = factor(distid), year = factor(year))
```

**Exercise:**

- Explore the difference between `lm()` and `feols()` to fit two-way fixed effects models that address our research question
- Compare results to a random effects model using `lmer()`
